---
title: "Kenya 2019 Census"
subtitle: "Summary analyses of the datasets"
author: "Levi Orero"
date: "6/12/2021"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
```

# Analysis

This section includes basic analysis to understand the data for each of the five datasets.

## Population data

```{r}
# load the required libraries
library(tidyverse)
library(sf)
library(sp)
library(rgdal)
library(rKenyaCensus)
library(janitor)
library(scales)
library(ggbeeswarm)
library(geomtextpath)
library(glue)
library(sysfonts)
library(showtext)
library(ggrepel)
```


```{r}
# import fonts
sysfonts::font_add_google("Roboto Slab","rs")
showtext::showtext_auto()
showtext_opts(dpi=300)
#sysfonts::font_add('Font Awesome 6 Brands', 'fonts/Font Awesome 6 Brands-Regular-400.otf')
```


```{r}
# read in the data
pop_county <- read_csv("data/pop_county.csv")
# Clean the names
pop_county <- clean_names(pop_county)
```



```{r}
# Arrange the top 10 counties by male population
gr_001 <- pop_county %>% arrange(-male) %>% slice(1:10) %>% 
  select(county, male)
gr_001
```

The counties with the highest male population were Nairobi, Kiambu, Nakuru, Kakamega, Bungoma, Meru, Machakos, Kilifi, Mombasa, Kisii.
```{r}
# write out the summary graph
fig1 <- 
gr_001 %>% 
  ggplot(aes(x = reorder(county,male),
             y = male)) +
  geom_bar(stat = "identity") +
  # flip coordinates for better readability
  coord_flip() +
  # add theme and labels
  labs(x = "County",
       y = "Population of males",
       caption = "Source: Kenya Population and Housing Census Results 2019") +
  theme_minimal() +
  # Change scale
  scale_y_continuous(labels = label_number(suffix = " M", scale = 1e-6))+ # millions
  # add font size
  theme(axis.text = element_text(size = 13),
        axis.title = element_text(size = 13),
        plot.caption = element_text(size = 12))
  fig1
  # write figure
ggsave(filename = "figures/males_population.jpg", width = 300, height = 250,
       units = "mm", dpi = 600)
```


```{r}
# Arrange the top 10 counties by female population
gr_002 <- 
pop_county %>% arrange(-female) %>% slice(1:10) %>% 
  select(county, female)
gr_002
```

The counties with the highest female population were Nairobi, Kiambu, Nakuru, Kakamega, Bungoma, Meru, Kilifi, Machakos, Kisii, Mombasa.

```{r}
# write out the summary graph
fig2 <- 
gr_002 %>% 
  ggplot(aes(x = reorder(county,female),
             y = female)) +
  geom_bar(stat = "identity", fill = "blue") +
  # flip coordinates for better readability
  coord_flip() +
  # add theme and labels
  labs(x = "County",
       y = "Population of females",
       caption = "Source: Kenya Population and Housing Census Results 2019") +
  theme_minimal() +
  # Change scale
  scale_y_continuous(labels = label_number(suffix = " M", scale = 1e-6)) + # millions
  # add font size
  theme(axis.text = element_text(size = 13),
        axis.title = element_text(size = 13),
        plot.caption = element_text(size = 12))
  fig2
  # write figure
ggsave(filename = "figures/females_population.jpg", width = 300, height = 250,
       units = "mm", dpi = 600)
```


```{r}
# Arrange the top 10 counties by intersex population
gr_003 <- pop_county %>% arrange(-intersex) %>% slice(1:10) %>% 
  select(county, intersex)
```

The counties with highest number of intersex individuals were Nairobi, Kiambu, Nakuru, Wajir, Meru, Kakamega, Kajiado, Kisii, Mandera and Bungoma.

```{r}
# import social
source('socials.R')
```


```{r}
# Plot a graph
# set caption
captionk = paste("Source: {Kenya Population and Housing Census Results 2019}<br>",
                 social_caption())
# write out the summary graph
fig3 <- 
gr_003 %>% 
  ggplot(aes(x = reorder(county,intersex),
             y = intersex)) +
  geom_bar(stat = "identity", fill = "brown") +
  # flip coordinates for better readability
  coord_flip() +
  # add theme and labels
  labs(x = "County",
       y = "Population of intersex individuals",
       caption = "Source: Kenya Population and Housing Census Results 2019") +
  theme_minimal() +
  #annotate("text", label = captionk)+
  # Change scale
  #scale_y_continuous(labels = label_number(suffix = " M", scale = 1e-6)) + # millions
  # add font size
  theme(axis.text = element_text(size = 13),
        axis.title = element_text(size = 13),
        plot.caption = element_text(size = 12))
  fig3
  # write figure
ggsave(filename = "figures/intersex_population.jpg", width = 300, height = 250,
       units = "mm", dpi = 600)
```


```{r}
# Arrange the top 10 counties by female:male ratio population
pop_county %>% mutate(ratio = round(female/male,2)) %>% 
  arrange(-ratio) %>% slice(1:10) %>% 
  select(county, ratio)
```

```{r}
# Arrange the bottom 10 counties by female:male ratio population
pop_county %>% mutate(ratio = round(female/male,2)) %>% 
  arrange(ratio) %>% slice(1:10) %>% 
  select(county, ratio)
```


## Households data

### Population

```{r}
# read in households data
households <- read_csv("data/households.csv")
```

```{r}
# Arrange the top 10 counties by population
households %>% arrange(-population) %>% slice(1:10)
```

The top 10 counties by population are Nairobi, Kiambu, Nakuru, Kakamega, Bungoma, Meru, Kilifi, Machakos, Kisii and Mombasa.

```{r}
# Arrange the bottom 10 counties by population
households %>% arrange(population) %>% slice(1:10)
```

The bottom ten counties by population are Lamu, Isiolo, Samburu, Tana River, Taita Taveta, Tharaka Nithi, Marsabit, Elgeyo-Marakwet, Laikipia and Vihiga.

### Household size

```{r}
# Arrange the top 10 counties by average household size
households %>% arrange(-average_household_size) %>% slice(1:10) %>% 
  select(county, average_household_size)
```

The top ten counties by average household size were Mandera (6.9), Wajir (6.1), Garissa (5.9), Marsabit (5.8), Turkana (5.6), West Pokot (5.3), Kwale (5.0), Kilifi (4.8), Narok (4.8), Samburu (4.7).

```{r}
# Arrange the bottom 10 counties by average household size
households %>% arrange(average_household_size) %>% slice(1:10) %>% 
  select(county, average_household_size)
```

The bottom 10 counties by average household size were Nairobi (2.9), Nyeri (3.0), Kirinyaga (3.0), Kiambu (3.0), Mombasa (3.1), Embu (3.3), Murang'a (3.3), Laikipia (3.4), Taita Taveta (3.5) and Machakos (3.5).

**Is average household size correlated with the population?**

```{r}
cor(households$population, households$average_household_size)
```
There is a weak negative correlation (-0.29) between county population and the average household size.

What is the average number of households for the counties with the lowest household sizes?

```{r}
households %>% arrange(average_household_size) %>% slice(1:10) %>% 
  select(county, average_household_size, number_of_households) %>% 
  summarize(mean = mean(number_of_households))

```
The mean number of is 428,180 households.

What is the average number of households for the counties with the highest household sizes?

```{r}
households %>% arrange(-average_household_size) %>% slice(1:10) %>% 
  select(county, average_household_size, number_of_households) %>% 
  summarize(mean = mean(number_of_households))
```
The mean number of households is 153,197.

We see that counties with high average household size have fewer households overall.

## Births data

### Total births

```{r}
# Top ten counties with most births in the last 5 years reference 2019
births_county %>% arrange(-total) %>%
  slice(1:10) %>% 
  select(county, total)
```


The counties with the most number of births in the past 5 years referenced 2019 were Nairobi, Kiambu, Nakuru, Kakamega, Bungoma, Meru, Kilifi, Mombasa, Narok and Kajiado.

```{r}
# Bottom counties with least births in the last 5 years reference 2019
births_county %>% arrange(total) %>%
  slice(1:10) %>% 
  select(county, total)
```

The counties with the least number of total births in the past 5 years referenced 2019 were Lamu, Isiolo, Samburu, Taita Taveta, Tana River, Tharaka Nith, Marsabit, Elgeyo Marakwet, vihiga and Laikipia.

Which counties have the least number of notified births?

```{r}
# Bottom counties with least births in the last 5 years reference 2019
births_county %>% arrange(percent_notified) %>%
  slice(1:10) %>% 
  select(county, percent_notified)
```

The counties with the least number of notified births were Wajir, Mandera, West Pokot, Marsabit, Turkana, Samburu, Garissa, Tana River, Isiolo and Narok.

## Assets Owned Data

### Stand alone radio
```{r}
# Top counties with radio percentage ownership
radio <- assets_owned %>% arrange(-stand_alone_radio) %>%
  slice(1:10) %>% 
  select(county, stand_alone_radio)
radio
```

Ten top counties with the highest owenership of stand alone radios were Nyeri, Nyandarua, Murang'a, Kirinyaga, Siaya, Kisumu, Laikipia, Machakos, Homa Bay and Embu.

If we compare this to internet and mobile devices/computers, we have the follwing:

```{r}
# compare radio, internet and mobile device use
assets_owned %>% 
  # internet, analogue tv and functional tv position
  mutate(ranking = rank(-internet),
         ranking_tv = rank(-analogue_tv),
         ranking_tv2 = rank(-functional_tv)) %>% 
  # filter radio use
  filter(county %in% radio$county) %>% 
  # select columns
  select(county, stand_alone_radio, desk_top_computer_laptop_tablet, internet, ranking,
         analogue_tv, ranking_tv, ranking_tv2) %>% 
  arrange(ranking)
```

Out of the ten counties, Nyeri, Kisumu, Machakos and Laikipia were the most highly ranked in terms of access to the internet. Therefore, they should be first in line to roll out the marketing of online goods to.

To verify this with goods bought online data:

```{r}
# read in goods online data
goods <- read_csv("data/goods_online.csv")
```
```{r}
# Add ranking based on goods search online
goods %>% 
  # add rank for number of online search
  mutate(ranking = rank(-searched_online_total)) %>% 
  # filter ten counties 
  filter(county %in% radio$county) %>% 
  # select columns
  select(county, searched_online_total, searched_online_total_perc, ranking) %>% 
  arrange(ranking)
```

Again, Machakos, Kisumu and Nyeri ranked top in terms of number of individuals who searched for goods online.

In Nyeri use radio, internet and functional tv
In Kisumu use radio, internet and analogue tv
In Machakos use radio, internet and analogue tv
