---
title: "E-commerce in Kenya"
subtitle: "How many Kenyans buy goods online?"
author: "Levi Orero"
date: "6/12/2021"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

# Analysis

This section includes basic analysis to understand the data for each of the five datasets.

## Households data

### Population
```{r}
# Arrange the top 10 counties by population
households %>% arrange(-population) %>% slice(1:10)
```

The top 10 counties by population are Nairobi, Kiambu, Nakuru, Kakamega, Bungoma, Meru, Kilifi, Macakos, Kisii and Mombasa.

```{r}
# Arrange the bottom 10 counties by population
households %>% arrange(population) %>% slice(1:10)
```

The bottom ten counties by population are Lamu, Isiolo, Samburu, Tana River, Taita Taveta, Tharaka Nithi, Marsabit, Elgeyo-Marakwet, Laikipia and Vihiga.

### Household size

```{r}
# Arrange the top 10 counties by average household size
households %>% arrange(-average_household_size) %>% slice(1:10) %>% 
  select(county, average_household_size)
```

The top ten counties by average household size were Mandera, Wajir, Garissa, Marsabit, Turkana, West Pokot, Kwale, Kilifi, Narok, Samburu.

```{r}
# Arrange the bottom 10 counties by average household size
households %>% arrange(average_household_size) %>% slice(1:10) %>% 
  select(county, average_household_size)
```

The bottom 10 counties by average household size were Nairobi, Nyeri, Kirinyaga, Kiambu, Mombasa, Embu, Murang'a, Laikipia, Taita Taveta and Machakos.

**Is average household size correlated with the population?**

```{r}
cor(households$population, households$average_household_size)
```
There is a weak negative correlation between population and the average household size.

What is the average number of households for the counties with the lowest household sizes?

```{r}
households %>% arrange(average_household_size) %>% slice(1:10) %>% 
  select(county, average_household_size, number_of_households) %>% 
  summarize(mean = mean(number_of_households))

```
The mean number of is 428,180 households.

What is the average number of households for the counties with the highest household sizes?

```{r}
households %>% arrange(-average_household_size) %>% slice(1:10) %>% 
  select(county, average_household_size, number_of_households) %>% 
  summarize(mean = mean(number_of_households))
```
The mean number of households is 153,197.

We see that counties with high average household size have fewer households overall.

## Births data

### Total births

```{r}
# Top ten counties with most births in the last 5 years reference 2019
births_county %>% arrange(-total) %>%
  slice(1:10) %>% 
  select(county, total)
```


The counties with the most number of births in the past 5 years referenced 2019 were Nairobi, Kiambu, Nakuru, Kakamega, Bungoma, Meru, Kilifi, Mombasa, Narok and Kajiado.

```{r}
# Bottom counties with least births in the last 5 years reference 2019
births_county %>% arrange(total) %>%
  slice(1:10) %>% 
  select(county, total)
```

The counties with the least number of total births in the past 5 years referenced 2019 were Lamu, Isiolo, Samburu, Taita Taveta, Tana River, Tharaka Nith, Marsabit, Elgeyo Marakwet, vihiga and Laikipia.

Which counties have the least number of notified births?

```{r}
# Bottom counties with least births in the last 5 years reference 2019
births_county %>% arrange(percent_notified) %>%
  slice(1:10) %>% 
  select(county, percent_notified)
```

The counties with the least number of notified births were Wajir, Mandera, West Pokot, marsabit, Turkana, Samburu, Garissa, Tana River, Isiolo and Narok.