---
title: "Kenya 2019 Census"
subtitle: "Summary analyses of the datasets"
author: "Levi Orero"
date: "6/12/2021"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
```

# Analysis

This section includes basic analysis to understand the data for each of the five datasets.

## Population data

```{r}
# load the required libraries
library(tidyverse)
library(sf)
library(sp)
library(rgdal)
library(rKenyaCensus)
library(janitor)
library(scales)
```

```{r}
# read in the data
pop_county <- read_csv("data/pop_county.csv")
# Clean the names
pop_county <- clean_names(pop_county)
```



```{r}
# Arrange the top 10 counties by male population
gr_001 <- pop_county %>% arrange(-male) %>% slice(1:10) %>% 
  select(county, male)
gr_001
```

The counties with the highest male population were Nairobi, Kiambu, Nakuru, Kakamega, Bungoma, Meru, Machakos, Kilifi, Mombasa, Kisii.
```{r}
# write out the summary graph
fig1 <- 
gr_001 %>% 
  ggplot(aes(x = reorder(county,male),
             y = male)) +
  geom_bar(stat = "identity") +
  # flip coordinates for better readability
  coord_flip() +
  # add theme and labels
  labs(x = "County",
       y = "Population of males",
       caption = "Source: Kenya Population and Housing Census Results 2019") +
  theme_minimal() +
  # Change scale
  scale_y_continuous(labels = label_number(suffix = " M", scale = 1e-6)) # millions
  fig1
  # write figure
ggsave(filename = "figures/males_population.png", width = 300, height = 250,
       units = "mm")
```


```{r}
# Arrange the top 10 counties by female population
gr_002 <- 
pop_county %>% arrange(-female) %>% slice(1:10) %>% 
  select(county, female)
gr_002
```

The counties with the highest female population were Nairobi, Kiambu, Nakuru, Kakamega, Bungoma, Meru, Kilifi, Machakos, Kisii, Mombasa.

```{r}
# write out the summary graph
fig2 <- 
gr_002 %>% 
  ggplot(aes(x = reorder(county,female),
             y = female)) +
  geom_bar(stat = "identity", fill = "blue") +
  # flip coordinates for better readability
  coord_flip() +
  # add theme and labels
  labs(x = "County",
       y = "Population of females",
       caption = "Source: Kenya Population and Housing Census Results 2019") +
  theme_minimal() +
  # Change scale
  scale_y_continuous(labels = label_number(suffix = " M", scale = 1e-6)) # millions
  fig2
  # write figure
ggsave(filename = "figures/females_population.png", width = 300, height = 250,
       units = "mm")
```


```{r}
# Arrange the top 10 counties by intersex population
pop_county %>% arrange(-intersex) %>% slice(1:10) %>% 
  select(county, intersex)
```

The counties with highest number of intersex individuals were Nairobi, Kiambu, Nakuru, Wajir, Meru, Kakamega, Kajiado, Kisii, Mandera and Bungoma.

```{r}
# Arrange the top 10 counties by female:male ratio population
pop_county %>% mutate(ratio = round(female/male,2)) %>% 
  arrange(-ratio) %>% slice(1:10) %>% 
  select(county, ratio)
```

```{r}
# Arrange the bottom 10 counties by female:male ratio population
pop_county %>% mutate(ratio = round(female/male,2)) %>% 
  arrange(ratio) %>% slice(1:10) %>% 
  select(county, ratio)
```


## Households data

### Population
```{r}
# Arrange the top 10 counties by population
households %>% arrange(-population) %>% slice(1:10)
```

The top 10 counties by population are Nairobi, Kiambu, Nakuru, Kakamega, Bungoma, Meru, Kilifi, Macakos, Kisii and Mombasa.

```{r}
# Arrange the bottom 10 counties by population
households %>% arrange(population) %>% slice(1:10)
```

The bottom ten counties by population are Lamu, Isiolo, Samburu, Tana River, Taita Taveta, Tharaka Nithi, Marsabit, Elgeyo-Marakwet, Laikipia and Vihiga.

### Household size

```{r}
# Arrange the top 10 counties by average household size
households %>% arrange(-average_household_size) %>% slice(1:10) %>% 
  select(county, average_household_size)
```

The top ten counties by average household size were Mandera, Wajir, Garissa, Marsabit, Turkana, West Pokot, Kwale, Kilifi, Narok, Samburu.

```{r}
# Arrange the bottom 10 counties by average household size
households %>% arrange(average_household_size) %>% slice(1:10) %>% 
  select(county, average_household_size)
```

The bottom 10 counties by average household size were Nairobi, Nyeri, Kirinyaga, Kiambu, Mombasa, Embu, Murang'a, Laikipia, Taita Taveta and Machakos.

**Is average household size correlated with the population?**

```{r}
cor(households$population, households$average_household_size)
```
There is a weak negative correlation between population and the average household size.

What is the average number of households for the counties with the lowest household sizes?

```{r}
households %>% arrange(average_household_size) %>% slice(1:10) %>% 
  select(county, average_household_size, number_of_households) %>% 
  summarize(mean = mean(number_of_households))

```
The mean number of is 428,180 households.

What is the average number of households for the counties with the highest household sizes?

```{r}
households %>% arrange(-average_household_size) %>% slice(1:10) %>% 
  select(county, average_household_size, number_of_households) %>% 
  summarize(mean = mean(number_of_households))
```
The mean number of households is 153,197.

We see that counties with high average household size have fewer households overall.

## Births data

### Total births

```{r}
# Top ten counties with most births in the last 5 years reference 2019
births_county %>% arrange(-total) %>%
  slice(1:10) %>% 
  select(county, total)
```


The counties with the most number of births in the past 5 years referenced 2019 were Nairobi, Kiambu, Nakuru, Kakamega, Bungoma, Meru, Kilifi, Mombasa, Narok and Kajiado.

```{r}
# Bottom counties with least births in the last 5 years reference 2019
births_county %>% arrange(total) %>%
  slice(1:10) %>% 
  select(county, total)
```

The counties with the least number of total births in the past 5 years referenced 2019 were Lamu, Isiolo, Samburu, Taita Taveta, Tana River, Tharaka Nith, Marsabit, Elgeyo Marakwet, vihiga and Laikipia.

Which counties have the least number of notified births?

```{r}
# Bottom counties with least births in the last 5 years reference 2019
births_county %>% arrange(percent_notified) %>%
  slice(1:10) %>% 
  select(county, percent_notified)
```

The counties with the least number of notified births were Wajir, Mandera, West Pokot, Marsabit, Turkana, Samburu, Garissa, Tana River, Isiolo and Narok.

## Assets Owned Data

### Stand alone radio
```{r}
# Top counties with radio percentage ownership
assets_owned %>% arrange(-stand_alone_radio) %>%
  slice(1:10) %>% 
  select(county, stand_alone_radio)
```


